Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-06-28 hazelnut-popl17.tex
Adding some todos, commentary in other areas of the action semantics. im starting to think of these theorems as semantic checksums. is that a good name?

\section{Hazelnut, Formally}
\label{sec:hazel}
\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}

@@

As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
every action in such a list, we couldn't hope to construct an ill-typed
term, so the statement here matches the mutually recursive structure of the
bidirectional typing judgements.

@@

\end{subequations}
\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}

\end{subequations}
\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.

@@

@@

\paragraph{Constructibility}

We give the following theorem to demonstrate that the rules for the action
$\aConstruct{\varphi}$ do indeed capture the process of building any term
of the form $\varphi$. Intuitively, its proof gives a list of composable
actions that, when applied one sequentially to an empty hole in focus,
produce any well-typed expression.

>>>>

\section{Hazelnut, Formally}
\label{sec:hazel}
\todo{revise text}\todo{add let and ifz}\todo{make eqn numbering work better}Hazelnut is based on the simply-typed lambda calculus extended with a single base type, $\tnum$. Its major constituents, introduced by example in the previous section, are:
\begin{itemize}
\item \textbf{H-types} and \textbf{H-expressions} (Sec. \ref{sec:holes}), which are terms with \emph{holes}. Holes mark subterms that are ``under construction.'' H-types classify H-expressions according to a {bidirectionally typed} static semantics.
\item \textbf{Z-types} and \textbf{Z-expressions} (Sec. \ref{sec:cursors}), which superimpose a single \emph{focus}\todo{cursor?} onto H-types and H-expressions (using Huet's \emph{zipper pattern} \cite{JFP::Huet1997}.)
\item \textbf{Actions} (Sec. \ref{sec:actions}), which move the focus or modify the subterm in focus.

@@

\begin{equation}\label{r:moveparent-hole}
  \inferrule{ }{
    \performMove{
      \hhole{\zwsel{\hexp}}
    }{
      \aMove{\dParent}
    }{
      \zwsel{\hhole{\hexp}}
    }
  }
\end{equation}
\end{subequations}

@@

As a consequence of applying Theorem \ref{thrm:actsafe} inductively at
every action in such a list, we couldn't hope to construct an ill-typed
term. Therefore the statement here matches the mutually recursive structure
of the bidirectional typing judgements.

@@

\subsubsection{Deletion} The $\aDel$ action replaces the selected subterm with an empty hole.

@@

Deletion within a Z-expression is similarly straightforward:
\begin{subequations}
\begin{equation}
  \inferrule{ }{
    \performSyn{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}{\tehole}
  }
\end{equation}
\begin{equation}
  \inferrule{ }{
    \performAna{\hGamma}{\zwsel{\hexp}}{\htau}{\aDel}{\zwsel{\hehole}}
  }
\end{equation}
%\end{subequations}
% The base case turns into a hole:
%\begin{subequations}
% \begin{equation}
% \inferrule{ }{
%   \performDel{\zwsel{\hexp}}{\hehole}
% }
% \end{equation}
% The rules for the recursive ascription case is shown below. The other recursive cases are analagous:
% \begin{equation}
%   \inferrule{
%     \performDel{\zexp}{\zexp'}
%   }{
%     \performDel{\zexp : \htau}{\zexp' : \htau}
%   }
% \end{equation}
% \begin{equation}
%   \inferrule{
%     \performTyp{\ztau}{\aDel}{\ztau'}
%   }{
%     \performDel{\hexp : \ztau}{\hexp : \ztau'}
%   }
% \end{equation}
\end{subequations}

\paragraph{Deletability}\todo{do we care?}
The meaning of the deletion action is to remove the entire term in focus,
and its definition does not depend on the structure of the term being
deleted. Therefore, there is no theorem to state that acts as a checksum on
the rules analogous to Theorem \ref{thrm:reach} or \ref{thrm:construct},
since we could not have forgotten any rules corresponding to the structure
of the terms.

\subsubsection{Construction} The construction actions, $\aConstruct{\varphi}$, are used to construct terms of a shape indicated by $\varphi$ into the program at or around the focus.

@@

\paragraph{Finishability}\todo{do we care?}

As with deletion, since the meaning of the finish action does not depend on
the structure of the term being finished other than requiring that it must
be one of the two hole forms, there is no checksum theorem to state for
this action.

@@

\paragraph{Constructibility}\todo{why is this with the zipper rules not the
    construct rules?}

We give the following theorem to demonstrate that the rules for the action
$\aConstruct{\varphi}$ do indeed capture the process of building any term
of the form $\varphi$. Intuitively, its proof gives a list of composable
actions that, when applied one sequentially to an empty hole in focus,
build any well-typed expression.
